Processing Script

### 2024-03-13 code/analysis-code/statistical-analysis.qmd
Additional Predictors
I used the codebook to find 5 more predictors for our outcome. I processed them and made sure there was no missingness. I did a little bit of EDA and created simple linear models with each of them.

---
title: "Statistical Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
```
## Load the cleaned data
```{r}
#path to data
data_location <- here("data","processed-data","processeddata2.rds")
#load data
mydata <- readRDS(data_location)
```

@@

# save fit results table  
table_file6 = here("results", "resulttable6.rds")
saveRDS(lmtable6, file = table_file6)
```

>>>>

---
title: "Statistical Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
```

## Load the cleaned data
```{r}
#path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

# save fit results table  
table_file6 = here("results", "resulttable6.rds")
saveRDS(lmtable6, file = table_file6)
```
### Seventh Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_currentlymarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit7 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_currentlymarried, mydata)

#generate clean output with estimates and p-values
tidy(lmfit7)

# save results from fit into a data frame 
lmtable7 <- tidy(lmfit7)

# save fit results table  
table_file7 = here("results", "resulttable7.rds")
saveRDS(lmtable7, file = table_file7)
```

# Eighth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_nevermarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit8 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_nevermarried, mydata)

#generate clean output with estimates and p-values
tidy(lmfit8)

# save results from fit into a data frame 
lmtable8 <- tidy(lmfit8)

# save fit results table  
table_file8 = here("results", "resulttable7.rds")
saveRDS(lmtable8, file = table_file8)
```

# Ninth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit9 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths, mydata)

#generate clean output with estimates and p-values
tidy(lmfit9)

# save results from fit into a data frame 
lmtable9 <- tidy(lmfit9)

# save fit results table  
table_file9 = here("results", "resulttable9.rds")
saveRDS(lmtable9, file = table_file9)
```

# Tenth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_upreg as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit10 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_upreg, mydata)

#generate clean output with estimates and p-values
tidy(lmfit10)

# save results from fit into a data frame 
lmtable10 <- tidy(lmfit10)

# save fit results table  
table_file10 = here("results", "resulttable10.rds")
saveRDS(lmtable10, file = table_file10)
```

# Eleventh Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_abs as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit11 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_abs, mydata)

#generate clean output with estimates and p-values
tidy(lmfit11)

# save results from fit into a data frame 
lmtable11 <- tidy(lmfit11)

# save fit results table  
table_file11 = here("results", "resulttable9.rds")
saveRDS(lmtable11, file = table_file11)
```

### 2024-03-13 code/eda-code/eda.qmd
Additional Predictors
I used the codebook to find 5 more predictors for our outcome. I processed them and made sure there was no missingness. I did a little bit of EDA and created simple linear models with each of them.

>>>>

We also need to explore the remaining predictor variables: `region`, `pct_currentlymarried`, `pct_nevermarried`, `rate_matdeaths`, `rate_matdeaths_upreg`, and `rate_matdeaths_abs`. Simple histograms will allow us to see the distribution of each variable and ensure there are no outliers. 
```{r}
# explore the region predictor
p22 <- mydata %>% ggplot(aes(x=region)) + 
  geom_bar() +
  ggtitle("Countries in Each Region") + 
  xlab("Region") + ylab("Number of Countries")
plot(p22)
figure_file = here("results", "figures", "exploratory-figures", "region_distribution.png")
ggsave(filename = figure_file, plot=p22)

# explore the current marriage predictor
p23 <- mydata %>% ggplot(aes(x=pct_currentlymarried)) + 
  geom_histogram() +
  ggtitle("Percent of Women Currently Married") + 
  xlab("Percentage") + ylab("Number of Countries")
plot(p23)
figure_file = here("results", "figures", "exploratory-figures", "married_distribution.png")
ggsave(filename = figure_file, plot=p23)

# explore the never married predictor
p24 <- mydata %>% ggplot(aes(x=pct_nevermarried)) + 
  geom_histogram() +
  ggtitle("Percent of Women Never Married") + 
  xlab("Percentage") + ylab("Number of Countries")
plot(p24)
figure_file = here("results", "figures", "exploratory-figures", "married_distribution.png")
ggsave(filename = figure_file, plot=p24)

# explore the total maternal mortality rate predictor
p25 <- mydata %>% ggplot(aes(x=rate_matdeaths)) + 
  geom_histogram() +
  ggtitle("Total Rate of Maternal Deaths") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p25)
figure_file = here("results", "figures", "exploratory-figures", "married_distribution.png")
ggsave(filename = figure_file, plot=p25)

# explore the maternal mortality rate from unintended pregnancies predictor
p26 <- mydata %>% ggplot(aes(x=rate_matdeaths_upreg)) + 
  geom_histogram() +
  ggtitle("Rate of Maternal Deaths from Unintended Pregnancies") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p26)
figure_file = here("results", "figures", "exploratory-figures", "married_distribution.png")
ggsave(filename = figure_file, plot=p26)

# explore outliers in maternal mortality from unintended pregnancies
mydata %>% filter(rate_matdeaths_upreg > 6000)

# explore the maternal mortality rate from abortions
p27 <- mydata %>% ggplot(aes(x=rate_matdeaths_abs)) + 
  geom_histogram() +
  ggtitle("Rate of Maternal Deaths from Abortions") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p27)
figure_file = here("results", "figures", "exploratory-figures", "married_distribution.png")
ggsave(filename = figure_file, plot=p27)

# explore outliers in maternal mortality from abortions
mydata %>% filter(rate_matdeaths_abs > 500)

```

P26: The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad and South Sudan that account for the highest rates of maternal mortality in this category.

P27: The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor.

### 2024-03-13 code/processing-code/processing-file.qmd
Additional Predictors
I used the codebook to find 5 more predictors for our outcome. I processed them and made sure there was no missingness. I did a little bit of EDA and created simple linear models with each of them.

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and birth control methods across different age groups. We will use three demographic variable: `country`, `region`, and `subregion`. The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable. We could also use `upreg_subreg` to explore unintended pregnancies by subregion. The predictor variables will include percentage of unintended pregnancies across three types of birth control: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). No method variables can serve as control groups for statistical tests. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, wra, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

Fortunately, the rest of the variables required for this analysis are complete including the `wra` variable with population size that we will need to standardize population counts later in the exploratory data analysis process. Each of the four variables describing the percentage of unintended pregnancies based on birth control type also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100)

# ensure mutations worked correctly
summary(rawdata2$pct_upreg_larcster)
summary(rawdata2$pct_upreg_sarc)
summary(rawdata2$pct_upreg_trad)
summary(rawdata2$pct_upreg_nouse)
```

@@

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larcster, total_sarc, total_trad, total_none)
```
After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larcster_usage = as.factor(case_when(
                                  total_larcster < 23.94865 ~ "Low",
                                  total_larcster >= 23.94865 & total_larcster < 1554.7609 ~ "Medium",
                                  total_larcster >= 1554.7609 ~ "High")))

>>>>

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and contraception methods. We will use three predictor variables: total number of women using the contraceptive method, region, number of safe abortions,  . The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable with additional options for outcome variables of total percentage of unintended pregnancies by each contraceptive method: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). We will use multiple predictor variables including: total number of women using the contraceptive method, region, subregion, percentage of married women, percentage of women never married, total maternal deaths per 100,000 live births, deaths from unintended pregnancy per 100,000 live births, and deaths related to abortion per 100000 abortions. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be summed to create total number of women using each type of contraception: 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, wra, region, subregion, pct_currentlymarried, pct_nevermarried, rate_matdeaths, rate_matdeaths_upreg, rate_matdeaths_abs, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

Fortunately, the rest of the variables required for this analysis are complete including the `wra` variable with population size that we will need to standardize population counts later in the exploratory data analysis process. Each of the four variables describing the percentage of unintended pregnancies based on birth control type and the two variables described percentage of married and non-married women also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100,
                     pct_currentlymarried = pct_currentlymarried * 100,
                     pct_nevermarried = pct_nevermarried * 100)

# ensure mutations worked correctly
summary(rawdata2$pct_upreg_larcster)
summary(rawdata2$pct_upreg_sarc)
summary(rawdata2$pct_upreg_trad)
summary(rawdata2$pct_upreg_nouse)
summary(rawdata2$pct_currentlymarried)
summary(rawdata2$pct_nevermarried)
```

@@

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larcster, total_sarc, total_trad, total_none)
```

After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larcster_usage = as.factor(case_when(
                                  total_larcster < 23.94865 ~ "Low",
                                  total_larcster >= 23.94865 & total_larcster < 1554.7609 ~ "Medium",
                                  total_larcster >= 1554.7609 ~ "High")))
